---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# characterize

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/characterize)](https://CRAN.R-project.org/package=characterize)
[![Codecov test coverage](https://codecov.io/gh/topepo/characterize/branch/main/graph/badge.svg)](https://app.codecov.io/gh/topepo/characterize?branch=main)
[![R-CMD-check](https://github.com/topepo/characterize/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/topepo/characterize/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of characterize is to report data driven characteristics of a model fit that describe the model. For example:

* the number of predictors used, 
* the number of terminal nodes, 

and so on. These are interesting characteristics of the model that often reflect complexity of the model.

## Installation

You can install the development version of characterize like so:

``` r
require(remotes)
remotes::install_github("topepo/characterize")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(characterize)
library(rpart)

cart_mod <- rpart(mpg ~ ., data = mtcars)
cart_mod

characterize(cart_mod)

# Also some low-level functions: 
.pluck_active_features(cart_mod) %>% 
  tidyr::unnest(value)
```

Objects with at least one method: `r object_list()`.

## Code of Conduct
  
Please note that the characterize project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
